---
title: "STAT 443 Project"
format: pdf
editor: visual
---

```{r}
library(forecast)
```

```{r}
# Function for variogram as defined in
# Bisgaard and Kulahci (2011). Time Series Analysis and Forecasting by Example, Wiley
variogram = function(y, lagmax=10, iprint=F)
{ G = rep(1,lagmax)
n = length(y)
if(lagmax>n) { lagmax = n-2 }
y1 = y[-1]; y2 = y[-n]
d1 = y1-y2; denom = var(d1)
for(k in 2:lagmax)
{ y1 = y[(k+1):n]; y2 = y[1:(n-k)]
dk = y1-y2
numer = var(dk)
G[k] = numer/denom
}
# H is the variogram assuming a stationary time series
H = rep(1,lagmax)
ac = c(acf(y,plot=F,lag.max=lagmax)$acf)
H = (1-ac[-1])/(1-ac[2])
if(iprint)
{ print(cbind(G,H))
mx = max(cbind(G,H))
matplot(1:lagmax,cbind(G,H),ylim=c(0,mx),ylab="variogram", xlab="lag")
abline(h=1)
}
list(G=G, H=H)
}
```

```{r}
persist_fc <- function(train, holdout, iprint = FALSE) {
  ntrain <- length(train)
  nh <- length(holdout)
  sse <- 0
  fc_vec <- numeric(nh)
  fc <- train[ntrain] 
  yt <- holdout[1]
  fc_vec[1] <- fc
  fcerror <- yt - fc
  sse <- sse + (fcerror^2)
  for (i in 2:nh) {
    yt <- holdout[i]
    fc <- holdout[i-1]
    fc_vec[i] <- fc
    fcerror <- yt - fc
    sse <- sse + (fcerror^2)
  }
  return(list(rmse = sqrt(sse/nh), fc = fc_vec))
}

# Linear in the previous observation
# b0,b1 from least squares or another estimation method
linreg_fc = function(train_last, holdout, b0,b1)
{ nh = length(holdout)
  previous = c(train_last, holdout[-nh])
  fc_vec = b0+b1*previous
  rmse = sqrt(mean((holdout-fc_vec)^2))
  return(list(rmse=rmse, fc=fc_vec))
}

# Simple Exponential Smoothing
esm_fc <- function(train, holdout, alpha, level, iprint = FALSE) {
  nh <- length(holdout)
  sse <- 0
  fc_vec <- numeric(nh)
  L <- level 
  for (i in 1:nh) {
    fc <- L
    fc_vec[i] <- fc
    yt <- holdout[i]
    fcerror <- yt - fc
    sse <- sse + (fcerror^2)
    L <- alpha * yt + (1 - alpha) * L
  }
  return(list(rmse = sqrt(sse / n_holdout), fc = fc_vec))
}
```

```{r}
plot_forecast <- function(holdout, fc, model_name = "Model") {
  
  x <- seq_along(holdout)
  
  plot(x, holdout, type = "l", col = "black", lwd = 2,
       main = paste("Forecast vs Actual -", model_name),
       ylab = "Value", xlab = "Time")
  
  lines(x, fc, col = "red", lwd = 2)
  
  legend("topleft",
         legend = c("Actual", "Forecast"),
         col = c("black", "red"),
         lty = 1, lwd = 2)
}
```

```{r}
arima_rolling_fc <- function(fitted_model, holdout, iprint = FALSE) {
  nh <- length(holdout)
  sse <- 0
  fc_vec <- numeric(nh)
  
  # Extract the training data from the fitted model
  train_data <- fitted_model$x
  current_data <- train_data
  
  for (i in 1:nh) {
    # Forecast one step ahead using current model
    fc <- forecast(Arima(current_data, model = fitted_model), h = 1)$mean[1]
    fc_vec[i] <- fc
    
    yt <- holdout[i]
    fcerror <- yt - fc
    sse <- sse + (fcerror^2)
    
    # Expand training data with actual observed value
    current_data <- c(current_data, yt)
  }
  
  return(list(rmse = sqrt(sse / nh), fc = fc_vec))
}
```


## Load Data

```{r}
# Load Spot Crude Oil Price
crude_oil <- read.csv("data/WTISPLC.csv")
crude_oil_ts <- ts(crude_oil$WTISPLC,start=c(1975,1,1),freq=12)

head(crude_oil_ts)
```


## Split Training - Holdout

```{r}
crude_oil_train_ts = window(crude_oil_ts, end = c(2015,12))
crude_oil_holdout_ts = window(crude_oil_ts, start = c(2016,1))

d_crude_oil_train_ts = diff(crude_oil_train_ts)
d_crude_oil_holdout_ts = diff(crude_oil_holdout_ts)

dl_crude_oil_train_ts = diff(log(crude_oil_train_ts))
dl_crude_oil_holdout_ts = diff(log(crude_oil_holdout_ts))
```

```{r}
end(d_crude_oil_holdout_ts)
```


```{r}
plot(crude_oil_ts)
```


# We first try ARIMA

## ARMA on difference

```{r}
dwti_arma = auto.arima(d_crude_oil_train_ts, seasonal=F, stationary=T)
print(dwti_arma)
```

### Rolling 1-step forward

```{r}
dwti_arma_rolling = arima_rolling_fc(dwti_arma, d_crude_oil_holdout_ts)
dwti_arma_rolling$rmse
```


## ARMA on log return

```{r}
dlwti_arma = auto.arima(dl_crude_oil_train_ts, seasonal=F, stationary=T)
print(dlwti_arma)
```
### Rolling 1-step forward

```{r}
dlwti_arma_rolling = arima_rolling_fc(dlwti_arma, dl_crude_oil_holdout_ts)
dlwti_arma_rolling$rmse
```

## Plots

```{r}
plot_forecast(d_crude_oil_holdout_ts, dwti_arma_rolling$fc, "Price Difference")
plot_forecast(dl_crude_oil_holdout_ts, dlwti_arma_rolling$fc, "Price Log Difference")
```




# We then introduce extrogenous variables

```{r}
usd_idx = read.csv("data/TWEXBGSMTH.csv")
ind_prod = read.csv("data/INDPRO.csv")

usd_idx$observation_date <- as.Date(usd_idx$observation_date)
ind_prod$observation_date <- as.Date(ind_prod$observation_date)

head(usd_idx)
```

```{r}
usd_ts <- ts(usd_idx$TWEXBGSMTH, start = c(1975,1), frequency = 12)
ind_ts <- ts(ind_prod$INDPRO, start = c(1975,1), frequency = 12)
```

```{r}
plot(usd_ts, main="U.S. Dollar Index", ylab="")
plot(ind_ts, main="Industrial Production Index", ylab="")
```

```{r}
d_usd_ts <- diff(usd_ts)
d_ind_ts <- diff(ind_ts)

dl_usd_ts <- diff(log(usd_ts))
dl_ind_ts <- diff(log(ind_ts))
```

## Split Train-Holdout

```{r}
ind_train_ts = window(ind_ts, end = c(2015,12))
ind_holdout_ts = window(ind_ts, start = c(2016,1))

d_ind_train_ts = window(d_ind_ts, end = c(2015,12))
d_ind_holdout_ts = window(d_ind_ts, start = c(2016,1), end=c(2026,1))

dl_ind_train_ts = window(dl_ind_ts, end = c(2015,12))
dl_ind_holdout_ts = window(dl_ind_ts, start = c(2016,1), end=c(2026,1))
```

```{r}

#BACKUPP
usd_train_ts = window(usd_ts, end = c(2015,12))
usd_holdout_ts = window(usd_ts, start = c(2016,1))

d_usd_train_ts = window(d_usd_ts, end = c(2015,12))
d_usd_holdout_ts = window(d_usd_ts, start = c(2016,1), end=c(2026,1))

dl_usd_train_ts = window(dl_usd_ts, end = c(2015,12))
dl_usd_holdout_ts = window(dl_usd_ts, start = c(2016,1), end=c(2026,1))
```


## Plot

```{r}
plot(d_usd_train_ts, main="Difference U.S. Dollar Index", ylab="")
plot(d_ind_train_ts, main="Difference Industrial Production Index", ylab="")

plot(dl_usd_train_ts, main="Diff. Log U.S. Dollar Index", ylab="")
plot(dl_ind_train_ts, main="Diff. Log Industrial Production Index", ylab="")
```

## Diagnostics: variogram, ACF, PACF








## Cross-correlation diagnostics for leading variables

```{r}
par(mfrow=c(1,2))

ccf(dl_ind_train_ts, dl_crude_oil_train_ts, main="CCF: dlog(INDPRO) vs dlog(WTI)")

plot(stats::lag(dl_ind_train_ts, -1), dl_ind_train_ts,
     xlab="dlog(INDPRO) lag 1", ylab="dlog(WTI)",
     main="Lag-1 scatter: INDPRO vs WTI")
```



















# Backup vvvvvvvvvvvvvvvvvvvvvvvvv


## We Introduce Exogenous Variable

```{r}
# Load Dow Jones Industrial Average Index Price
djia <- read.csv("data/DJIA.csv")
djia_ts <- ts(djia$DJIA, start=c(2016,2,8), freq=365)
head(djia_ts)
```

```{r}
plot(djia_ts)
```

```{r}
djia_price = djia$DJIA
d_djia = diff(djia_price)
dl_djia = diff(log(djia_price))

par(mfrow=c(2,2))
plot.ts(djia_price)
plot.ts(dl_djia)
acf(dl_djia, na.action = na.pass)
pacf(dl_djia, na.action = na.pass)
```

```{r}
variogram(djia_price, lagmax=20, iprint=T)
variogram(dl_djia, lagmax=20, iprint=T)
```

```{r}
djia_arma = auto.arima(dl_djia, seasonal=F, stationary=T)
print(djia_arma)
```










fesrgdrjsdxfgsrtjklhnsegbr


```{r}
par(mfrow=c(2,2))
plot.ts(oil_price)
plot.ts(d_wti)
acf(d_wti)
pacf(d_wti)
```

## Data

```{r}
oil_price = crude_oil$WTISPLC
d_wti = diff(oil_price)
dl_wti = diff(log(oil_price))
```

## Simple Rules

```{r}

```

```{r}
plot.ts(dl_wti)
acf(dl_wti)
pacf(dl_wti)
```

```{r}
variogram(oil_price, lagmax=20, iprint=T)
variogram(d_wti, lagmax=20, iprint=T)
variogram(dl_wti, lagmax=20, iprint=T)
```