---
title: "STAT 443 Project"
format: pdf
editor: visual
---

```{r}
library(forecast)
```

```{r}
# Function for variogram as defined in
# Bisgaard and Kulahci (2011). Time Series Analysis and Forecasting by Example, Wiley
variogram = function(y, lagmax=10, iprint=F)
{ G = rep(1,lagmax)
n = length(y)
if(lagmax>n) { lagmax = n-2 }
y1 = y[-1]; y2 = y[-n]
d1 = y1-y2; denom = var(d1)
for(k in 2:lagmax)
{ y1 = y[(k+1):n]; y2 = y[1:(n-k)]
dk = y1-y2
numer = var(dk)
G[k] = numer/denom
}
# H is the variogram assuming a stationary time series
H = rep(1,lagmax)
ac = c(acf(y,plot=F,lag.max=lagmax)$acf)
H = (1-ac[-1])/(1-ac[2])
if(iprint)
{ print(cbind(G,H))
mx = max(cbind(G,H))
matplot(1:lagmax,cbind(G,H),ylim=c(0,mx),ylab="variogram", xlab="lag")
abline(h=1)
}
list(G=G, H=H)
}
```

```{r}
persist_fc <- function(train, holdout, iprint = FALSE) {
  ntrain <- length(train)
  nh <- length(holdout)
  sse <- 0
  fc_vec <- numeric(nh)
  fc <- train[ntrain] 
  yt <- holdout[1]
  fc_vec[1] <- fc
  fcerror <- yt - fc
  sse <- sse + (fcerror^2)
  for (i in 2:nh) {
    yt <- holdout[i]
    fc <- holdout[i-1]
    fc_vec[i] <- fc
    fcerror <- yt - fc
    sse <- sse + (fcerror^2)
  }
  return(list(rmse = sqrt(sse/nh), fc = fc_vec))
}

# Linear in the previous observation
# b0,b1 from least squares or another estimation method
linreg_fc = function(train_last, holdout, b0,b1)
{ nh = length(holdout)
  previous = c(train_last, holdout[-nh])
  fc_vec = b0+b1*previous
  rmse = sqrt(mean((holdout-fc_vec)^2))
  return(list(rmse=rmse, fc=fc_vec))
}

# Simple Exponential Smoothing
esm_fc <- function(train, holdout, alpha, level, iprint = FALSE) {
  nh <- length(holdout)
  sse <- 0
  fc_vec <- numeric(nh)
  L <- level 
  for (i in 1:nh) {
    fc <- L
    fc_vec[i] <- fc
    yt <- holdout[i]
    fcerror <- yt - fc
    sse <- sse + (fcerror^2)
    L <- alpha * yt + (1 - alpha) * L
  }
  return(list(rmse = sqrt(sse / nh), fc = fc_vec))
}
```

```{r}
plot_forecast <- function(holdout, fc, model_name = "Model") {
  
  x <- seq_along(holdout)
  
  plot(x, holdout, type = "l", col = "black", lwd = 2,
       main = paste("Forecast vs Actual -", model_name),
       ylab = "Value", xlab = "Time")
  
  lines(x, fc, col = "red", lwd = 2)
  
  legend("topleft",
         legend = c("Actual", "Forecast"),
         col = c("black", "red"),
         lty = 1, lwd = 2)
}
```

```{r}
arima_rolling_fc <- function(fitted_model, holdout, iprint = FALSE) {
  nh <- length(holdout)
  sse <- 0
  fc_vec <- numeric(nh)
  
  # Extract the training data from the fitted model
  train_data <- fitted_model$x
  current_data <- train_data
  
  for (i in 1:nh) {
    # Forecast one step ahead using current model
    fc <- forecast(Arima(current_data, model = fitted_model), h = 1)$mean[1]
    fc_vec[i] <- fc
    
    yt <- holdout[i]
    fcerror <- yt - fc
    sse <- sse + (fcerror^2)
    
    # Expand training data with actual observed value
    current_data <- c(current_data, yt)
  }
  
  return(list(rmse = sqrt(sse / nh), fc = fc_vec))
}

rolling_armax_fixed <- function(y_train, x_train, y_holdout, x_holdout,
                                order, include_mean = TRUE) {
  nh <- length(y_holdout)
  fc_vec <- numeric(nh)

  y_hist <- y_train
  x_hist <- as.matrix(x_train)
  x_holdout <- as.matrix(x_holdout)

  for (i in 1:nh) {
    fit <- forecast::Arima(
      y_hist,
      order = order,
      xreg = x_hist,
      include.mean = include_mean,
      method = "ML"
    )

    newx <- matrix(x_holdout[i, ], nrow = 1)

    fc_vec[i] <- forecast::forecast(
      fit,
      xreg = newx,
      h = 1
    )$mean[1]

    y_hist <- c(y_hist, y_holdout[i])
    x_hist <- rbind(x_hist, newx)
  }

  rmse <- sqrt(mean((y_holdout - fc_vec)^2))
  list(rmse = rmse, fc = fc_vec)
}
```

```{r}
rolling_armax <- function(y_train, x_train, y_holdout, x_holdout,
                          order = c(0,0,1), include_mean = TRUE) {
  nh <- length(y_holdout)
  fc_vec <- numeric(nh)

  y_hist <- as.numeric(y_train)
  x_hist <- as.matrix(x_train)
  x_holdout <- as.matrix(x_holdout)

  for (i in 1:nh) {
    fit <- forecast::Arima(
      y_hist,
      order = order,
      xreg = x_hist,
      include.mean = include_mean,
      method = "ML"
    )

    newx <- matrix(x_holdout[i, ], nrow = 1)

    fc_vec[i] <- as.numeric(
      forecast::forecast(fit, xreg = newx, h = 1)$mean
    )

    # update history with the ACTUAL next observation
    y_hist <- c(y_hist, y_holdout[i])
    x_hist <- rbind(x_hist, newx)
  }

  rmse <- sqrt(mean((y_holdout - fc_vec)^2))
  list(rmse = rmse, fc = fc_vec)
}

armax_fc = function(tsdata, ntrain, order, traincoef, xreg)
{
  coef_names <- names(traincoef)
  
  include_mean <- any(coef_names %in% c("intercept", "mean"))
  
  obj = arima(tsdata,
              order = order,
              init = traincoef,
              fixed = traincoef,
              xreg = xreg,
              include.mean = include_mean,
              method = "ML",
              optim.control = list(maxit = 0))
  
  fc = tsdata - obj$residuals
  
  holdout_fc = fc[(ntrain+1):length(tsdata)]
  holdout = tsdata[(ntrain+1):length(tsdata)]
  
  rmse = sqrt(mean((holdout - holdout_fc)^2))
  
  return(list(rmse = rmse, fc = holdout_fc))
}
```

## Load Data

```{r}
# Load Spot Crude Oil Price
#crude_oil <- read.csv("data/WTISPLC.csv")
#crude_oil_ts <- ts(crude_oil$WTISPLC,start=c(1975,1,1),freq=12)

#head(crude_oil_ts)

# Load Spot Crude Oil Price
crude_oil <- read.csv("data/bloomberg_oil_spot_monthly.tsv", header = TRUE, sep = "\t")
crude_oil$Date <- as.Date(crude_oil$Date)
crude_oil <- crude_oil[order(crude_oil$Date), ]

crude_oil_ts <- ts(
  crude_oil$Last.Price,
  start = c(as.numeric(format(min(crude_oil$Date), "%Y")),
            as.numeric(format(min(crude_oil$Date), "%m"))),
  frequency = 12
)

head(crude_oil_ts)
#write.table(crude_oil_ts, "savedData/crude_oil_ts_new.csv", sep = ",", row.names = FALSE)
```

## Split Training - Holdout

```{r}
crude_oil_train_ts = window(crude_oil_ts, start = c(1975,1), end = c(2010,12))
crude_oil_holdout_ts = window(crude_oil_ts, start = c(2011,1), end = c(2026,1))

d_crude_oil_train_ts = diff(crude_oil_train_ts)
d_crude_oil_holdout_ts = diff(crude_oil_holdout_ts)

dl_crude_oil_ts <- diff(log(crude_oil_ts))
dl_crude_oil_train_ts = diff(log(crude_oil_train_ts))
dl_crude_oil_holdout_ts = diff(log(crude_oil_holdout_ts))
```

```{r}
end(d_crude_oil_holdout_ts)
```

##Diagnostic Plots

```{r}
plot(crude_oil_ts)
```

```{r}
par(mfrow = c(2,3))

plot.ts(crude_oil_train_ts)
acf(crude_oil_train_ts)
pacf(crude_oil_train_ts)

plot.ts(d_crude_oil_train_ts)
acf(d_crude_oil_train_ts)
pacf(d_crude_oil_train_ts)

plot.ts(dl_crude_oil_train_ts)
acf(dl_crude_oil_train_ts)
pacf(dl_crude_oil_train_ts)
```

```{r}
variogram(crude_oil_train_ts, lagmax=25, iprint=T)
```

```{r}
variogram(d_crude_oil_train_ts, lagmax=25, iprint=T)
```

```{r}
variogram(dl_crude_oil_train_ts, lagmax=25, iprint=T)
```

# We first try ARIMA

## ARMA on difference

```{r}
dwti_arma = auto.arima(d_crude_oil_train_ts, seasonal=F, stationary=T)
print(dwti_arma)
```

### Rolling 1-step forward

```{r}
dwti_arma_rolling = arima_rolling_fc(dwti_arma, d_crude_oil_holdout_ts)
dwti_arma_rolling$rmse
```

## ARMA on log return

```{r}
dlwti_arma = auto.arima(dl_crude_oil_train_ts, seasonal=F, stationary=T)
print(dlwti_arma)
```

### Rolling 1-step forward

```{r}
dlwti_arma_rolling = arima_rolling_fc(dlwti_arma, dl_crude_oil_holdout_ts)
dlwti_arma_rolling$rmse
```

## Plots

```{r}
plot_forecast(d_crude_oil_holdout_ts, dwti_arma_rolling$fc, "Price Difference")
plot_forecast(dl_crude_oil_holdout_ts, dlwti_arma_rolling$fc, "Price Log Difference")
```

# We then introduce extrogenous variables and split Train-Holdout

```{r}
#usd_idx = read.csv("data/TWEXBGSMTH.csv")
#usd_idx = read.csv("data/dxy_monthly.csv")
#ind_prod = read.csv("data/INDPRO.csv")
#opec_prod = read.csv("data/opcrtotl_monthly.tsv")

#usd index
usd_idx <- read.csv("data/dxy_monthly.tsv", header = TRUE, sep = "\t")
usd_idx$Date <- as.Date(usd_idx$Date)
usd_idx <- usd_idx[order(usd_idx$Date), ]

usd_idx_ts <- ts(
  usd_idx$Last.Price,
  start = c(1970,04),
  end = c(2026,04),
  frequency = 12
)
usd_idx_train_ts = window(usd_idx_ts, start = c(1975,1), end = c(2010,12))
usd_idx_holdout_ts = window(usd_idx_ts, start = c(2011,1), end = c(2026,1))

#Industrial production
ind_prod = read.csv("data/INDPRO.csv")
ind_prod$observation_date <- as.Date(ind_prod$observation_date)
ind_prod_ts <- ts(ind_prod$INDPRO, start = c(1919,1), end = c(2026,2), frequency = 12)
ind_prod_train_ts = window(ind_prod_ts, start = c(1975,1), end = c(2010,12))
ind_prod_holdout_ts = window(ind_prod_ts, start = c(2011,1), end = c(2026,1))

# OPEC oil production
opec_prod <- read.csv("data/opcrtotl_monthly.tsv", header = TRUE, sep = "\t")
opec_prod$Date <- as.Date(opec_prod$Date)
opec_prod <- opec_prod[order(opec_prod$Date), ]

opec_prod_ts <- ts(
  opec_prod$Last.Price,
  start = c(1970,12),
  end = c(2026,03),
  frequency = 12
)
opec_prod_train_ts = window(opec_prod_ts, start = c(1975,1), end = c(2010,12))
opec_prod_holdout_ts = window(opec_prod_ts, start = c(2011,1), end = c(2026,1))

# World MSCI Oil/Energy Company Index
#TO DO 



head(usd_idx)

#Saving data
#write.table(usd_idx_train_ts, "savedData/usd_idx_train_ts_new.csv", sep = ",", row.names = FALSE)
#write.table(usd_idx_holdout_ts, "savedData/usd_idx_holdout_ts_new.csv", sep = ",", row.names = FALSE)

#write.table(ind_prod_train_ts, "savedData/ind_prod_train_ts_new.csv", sep = ",", row.names = FALSE)
#write.table(ind_prod_holdout_ts, "savedData/ind_prod_holdout_ts_new.csv", sep = ",", row.names = FALSE)

#write.table(opec_prod_train_ts, "savedData/opec_prod_train_ts_new.csv", sep = ",", row.names = FALSE)
#write.table(opec_prod_holdout_ts, "savedData/opec_prod_holdout_ts_new.csv", sep = ",", row.names = FALSE)
```

```{r}
#usd_ts <- ts(usd_idx$Last.Price, start = c(1975,1), frequency = 12)
#ind_ts <- ts(ind_prod$INDPRO, start = c(1975,1), frequency = 12)
#opec_ts <- ts(opec_prod$Last.Price, start = c(1975,1), frequency = 12)
```

```{r}
plot(usd_idx_ts, main="U.S. Dollar Index", ylab="")
plot(ind_prod_ts, main="Industrial Production Index", ylab="")
plot(opec_prod_ts, main="OPEC Oil Production Index", ylab="")
```

### Transform log-differencing

```{r}
d_usd_ts <- diff(usd_idx_ts)
dl_usd_idx_ts <- diff(log(usd_idx_ts))
dl_usd_idx_train_ts <- diff(log(usd_idx_train_ts))

#write.table(d_usd_ts, "savedData/d_usd_ts_new.csv", sep = ",", row.names = FALSE)
#write.table(dl_usd_idx_train_ts, "savedData/dl_usd_idx_train_ts_new.csv", sep = ",", row.names = FALSE)


d_ind_ts <- diff(ind_prod_ts)
dl_ind_ts <- diff(log(ind_prod_ts))
dl_ind_train_ts <- diff(log(ind_prod_train_ts))

#write.table(d_ind_ts, "savedData/d_ind_ts_new.csv", sep = ",", row.names = FALSE)
#write.table(dl_ind_train_ts, "savedData/dl_ind_train_ts_new.csv", sep = ",", row.names = FALSE)


d_opec_ts <- diff(opec_prod_ts)
dl_opec_prod_ts <- diff(log(opec_prod_ts))
dl_opec_idx_train_ts <- diff(log(opec_prod_train_ts))

#write.table(d_opec_ts, "savedData/d_opec_ts_new.csv", sep = ",", row.names = FALSE)
#write.table(dl_opec_idx_train_ts, "savedData/dl_opec_idx_train_ts_new.csv", sep = ",", row.names = FALSE)

```

## Plot

```{r}
#plot(dl_usd_idx_train_ts, main="Difference U.S. Dollar Index", ylab="")
plot(dl_usd_idx_train_ts, main="Diff. Log U.S. Dollar Index", ylab="")

#plot(d_ind_train_ts, main="Difference Industrial Production Index", ylab="")
plot(dl_ind_train_ts, main="Diff. Log Industrial Production Index", ylab="")

#plot(d_opec_idx_train_ts, main="Difference Industrial Production Index", ylab="")
plot(dl_opec_idx_train_ts, main="Diff. Log OPEC Oil Production Index", ylab="")
```

## Diagnostics: variogram, ACF, PACF

```{r}
par(mfrow=c(1,2))

acf(dl_crude_oil_train_ts, 
    main = "ACF: dlog(crude_oil_train)")

pacf(dl_crude_oil_train_ts, 
     main = "PACF: dlog(crude_oil_train)")

```

## ARMAX

### Cross-correlation diagnostics for leading variables

```{r}

par(mfrow=c(1,2))

ccf(dl_ind_train_ts, dl_crude_oil_train_ts, main="CCF: dlog(INDPRO) vs dlog(crude_oil_train)")

plot(stats::lag(dl_ind_train_ts, -1), dl_ind_train_ts,
     xlab="dlog(INDPRO) lag 1", ylab="dlog(crude_oil_train)",
     main="Lag-1 scatter: INDPRO vs crude_oil_train")
```

```{r}
par(mfrow=c(1,2))

ccf(dl_opec_idx_train_ts, dl_crude_oil_train_ts, main="CCF: dlog(OPEC) vs dlog(Bloomberg)")

plot(stats::lag(dl_opec_idx_train_ts, -1), dl_opec_idx_train_ts,
     xlab="dlog(OPEC) lag 1", ylab="dlog(Bloomberg)",
     main="Lag-1 scatter: OPEC vs Bloomberg")
```

```{r}
par(mfrow=c(1,2))

ccf(dl_usd_idx_train_ts, dl_crude_oil_train_ts, main="CCF: dlog(DXY) vs dlog(Bloomberg)")

plot(stats::lag(dl_usd_idx_train_ts, -1), dl_usd_idx_train_ts,
     xlab="dlog(DXY) lag 1", ylab="dlog(Bloomberg)",
     main="Lag-1 scatter: DXY vs Bloomberg")
```

### Align data

```{r}
df_all <- ts.intersect(
  oil  = dl_crude_oil_ts,
  ind  = dl_ind_ts,
  opec = dl_opec_prod_ts,
  usd  = dl_usd_idx_ts
)

start(df_all)
end(df_all)
```

### Split Train/Holdout

```{r, warning=FALSE}
train_all <- window(df_all, start = c(1975, 2), end = c(2010, 12))
holdout_all <- window(df_all, start = c(2011, 1), end = c(2026, 1))

y_train <- train_all[, "oil"]
y_holdout <- holdout_all[, "oil"]

xreg_train <- as.matrix(train_all[, c("ind", "opec", "usd")])
xreg_holdout <- as.matrix(holdout_all[, c("ind", "opec", "usd")])
```

### ARIMA baseline

```{r, warning=FALSE}
# ARIMA baseline
arima_model <- auto.arima(
  y_train,
  seasonal = FALSE,
  stationary = TRUE
)

print(arima_model)

arima_res <- arima_rolling_fc(arima_model, y_holdout)
```

### ARMAX candidate model

```{r, warning=FALSE}
armax_model <- auto.arima(
  y_train,
  xreg = xreg_train,
  seasonal = FALSE,
  stationary = TRUE
)

print(armax_model)
```

### extract nonseasonal ARMA order

```{r}
order_armax <- c(armax_model$arma[1], 0, armax_model$arma[2])
include_mean_armax <- any(names(armax_model$coef) %in% c("intercept", "mean"))
```


### Rolling 1-step ARMAX

```{r, warning=FALSE}
armax_res <- rolling_armax_fixed(
  y_train = y_train,
  x_train = xreg_train,
  y_holdout = y_holdout,
  x_holdout = xreg_holdout,
  order = order_armax,
  include_mean = include_mean_armax
)
```

### RMSE comparison

```{r, warning=FALSE}
cat("ARIMA RMSE:", arima_res$rmse, "\n")
cat("ARMAX RMSE:", armax_res$rmse, "\n")
```


### Plot forecasts

```{r}
plot(as.numeric(y_holdout), type = "l", col = "black", lwd = 2,
     main = "Rolling 1-step Forecast Comparison",
     ylab = "dlog(crude oil)", xlab = "Time")

lines(as.numeric(arima_res$fc), col = "blue", lwd = 2)
lines(as.numeric(armax_res$fc), col = "red", lwd = 2)

legend("topleft",
       legend = c("Actual", "ARIMA", "ARMAX"),
       col = c("black", "blue", "red"),
       lty = 1, lwd = 2)
```


## Simple Rules

# TODO